\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-07-22 analysis/paper/paper.tex
fix hazard ratio plot

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(
R+   exposure = c(0, 1),
R+   futime = seq(0, 1000,
R+     length.out = 100
R+   )
R+ )
R> 
R> # Set the offset to zero
R> hazard_data$offset <- 0
R> # Use predict to get the fitted values, and exponentiate to
R> # transform to the right scale
R> hazard_data$hazard <- exp(predict(fit4,
R+   newdata = hazard_data,
R+   type = "link"
R+ ))
R> # Add labels for plots
R> hazard_data$Status <- factor(hazard_data$exposure,
R+   labels = c("NoTrans", "Trans")
R+ )
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R+   geom_line() +
R+   paper_gg_theme +  
R+   ylab("Hazard") +
R+   xlab("Follow-up time")
\end{CodeInput}

@@

In this article, we presented the \proglang{R} package \pkg{casebase},
which provides functions for fitting smooth parametric hazards and
estimating CIFs using case-base sampling. We outlined the theoretical
underpinnings of the approach, we provided details about our
implementation, and we illustrated the merits of the approach and the
package through four case studies.

As a methodological framework, case-base sampling is very flexible. This
flexibility has been explored before in the literature: for example,
Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
to model a time-dependent exposure variable in a vaccine safety study.
As another example, Saarela and Arjas \citeyearpar{saarela2015non}
combined case-base sampling and a Bayesian non-parametric framework to
compute individualized risk assessments for chronic diseases. In the
case studies above, we explored this flexibility along two fronts. On
the one hand, we showed how splines could be used as part of the linear
predictor to model the effect of time on the hazard. This strategy
yielded estimates of the survival function that were qualitatively
similar to semiparametric estimates derived from Cox regression;
however, case-base sampling led to estimates of the survival function
that \emph{vary smoothly in time}. On the other hand, we also displayed
the flexibility of case-base sampling by showing how it could be
combined with penalized logistic regression to perform variable
selection. Even though we did not illustrate it in this article,
case-base sampling can also be combined with the framework of
\emph{generalized additive models}. This functionality has been
implemented in our package \pkg{casebase}. Similarly, case-base sampling
can also be combined with quasi-likelihood estimation to fit survival
models that can account for the presence of over-dispersion. All of
these examples illustrate how the case-base sampling framework in
general, and the package \pkg{casebase} in particular, allows the user
to fit a broad and flexible family of survival functions.

@@

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \pkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
that capture both the event type and time. Both approaches to modeling
yield user-friendly code. However, in terms of output, both approaches
differ significantly. Case-base sampling produces smooth hazards and
smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
models produce step-wise cumulative incidence functions and never
explicitly model the hazard function. Qualitatively, we showed that by
using splines in the linear predictor, all three models yielded similar
curves. However, the smooth nature of the output of \pkg{casebase}
provides a more intuitive interpretation for consumers of these
predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
comparison between the Cox model and case-base sampling.

@@

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
want to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools would then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
Again, we hope to include any resulting methodology as part of this
package.

@@

This report was generated on 2020-07-21 19:46:24 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.0 (2020-04-24)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.5
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.5.6               dotwhisker_0.5.0         
#>  [3] timereg_1.9.4             glue_1.4.1               
#>  [5] scales_1.1.0              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_0.8.5              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.0.2               ggplot2_3.3.0            
#> [13] tidyverse_1.3.0           lubridate_1.7.8          
#> [15] visreg_2.6.1              tibble_3.0.1             
#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.1-12          
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.1.0          glmnet_3.0-2             
#> [25] Matrix_1.2-18             data.table_1.12.8        
#> [27] cowplot_1.0.0             colorspace_1.4-1         
#> [29] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
#>  [7] fs_1.4.1            rstudioapi_0.11     MatrixModels_0.4-1 
#> [10] fansi_0.4.1         mvtnorm_1.1-0       xml2_1.3.2         
#> [13] codetools_0.2-16    knitr_1.29          Formula_1.2-3      
#> [16] jsonlite_1.6.1      cluster_2.1.0       dbplyr_1.4.3       
#> [19] png_0.1-7           compiler_4.0.0      httr_1.4.1         
#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.55      
#> [28] tools_4.0.0         gtable_0.3.0        Rcpp_1.0.4.6       
#> [31] cellranger_1.1.0    vctrs_0.3.1         nlme_3.1-147       
#> [34] iterators_1.0.12    xfun_0.15           rvest_0.3.5        
#> [37] lifecycle_0.2.0     polspline_1.1.17    muhaz_1.2.6.1      
#> [40] MASS_7.3-51.5       zoo_1.8-7           hms_0.5.3          
#> [43] sandwich_2.5-1      SparseM_1.78        RColorBrewer_1.1-2 
#> [46] rticles_0.14        yaml_2.2.1          gridExtra_2.3      
#> [49] rms_5.1-4           rpart_4.1-15        latticeExtra_0.6-29
#> [52] stringi_1.4.6       foreach_1.5.0       checkmate_2.0.0    
#> [55] lava_1.6.7          shape_1.4.4         rlang_0.4.6        
#> [58] pkgconfig_2.0.3     evaluate_0.14       lattice_0.20-41    
#> [61] htmlwidgets_1.5.1   cmprsk_2.2-9        tidyselect_1.1.0   
#> [64] deSolve_1.28        R6_2.4.1            generics_0.0.2     
#> [67] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
#> [70] withr_2.2.0         pillar_1.4.4        haven_2.2.0        
#> [73] foreign_0.8-78      mgcv_1.8-31         nnet_7.3-13        
#> [76] mstate_0.2.12       modelr_0.1.6        crayon_1.3.4       
#> [79] rmarkdown_2.1       jpeg_0.1-8.1        grid_4.0.0         
#> [82] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
#> [85] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.0       
#> [88] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}

>>>>

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(
R+   exposure = c(0, 1),
R+   futime = seq(0, 425,
R+     length.out = 100
R+   )
R+ )
R> 
R> # Set the offset to zero
R> hazard_data$offset <- 0
R> # Use predict to get the fitted values, and exponentiate to
R> # transform to the right scale
R> hazard_data$hazard <- exp(predict(fit4,
R+   newdata = hazard_data,
R+   type = "link"
R+ ))
R> # Add labels for plots
R> hazard_data$Status <- factor(hazard_data$exposure,
R+   labels = c("NoTrans", "Trans")
R+ )
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R+   geom_line() +
R+   paper_gg_theme +  
R+   ylab("Hazard") +
R+   xlab("Follow-up time")
\end{CodeInput}

@@

In this article, we presented the \proglang{R} package \pkg{casebase},
which provides functions for fitting smooth parametric hazards and
estimating CIFs using case-base sampling. Our package also provide
several functions to produce graphical summaries of the data and the
results. We outlined the theoretical underpinnings of the approach, we
provided details about our implementation, and we illustrated the merits
of the approach and the package through four case studies.

As a methodological framework, case-base sampling is very flexible. This
flexibility has been explored before in the literature: for example,
Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
to model a time-dependent exposure variable in a vaccine safety study.
As another example, Saarela and Arjas \citeyearpar{saarela2015non}
combined case-base sampling and a Bayesian non-parametric framework to
compute individualized risk assessments for chronic diseases. In the
case studies above, we further explored this flexibility along two
fronts. On the one hand, we showed how splines could be used as part of
the linear predictor to model the effect of time on the hazard. This
strategy yielded estimates of the survival function that were
qualitatively similar to semiparametric estimates derived from Cox
regression; however, case-base sampling led to estimates of the survival
function that \emph{vary smoothly in time} and are thus easier to
interpret. On the other hand, we also displayed the flexibility of
case-base sampling by showing how it could be combined with penalized
logistic regression to perform variable selection. Even though we did
not illustrate it in this article, case-base sampling can also be
combined with the framework of \emph{generalized additive models}. This
functionality has already been implemented in the package. Similarly,
case-base sampling can be combined with quasi-likelihood estimation to
fit survival models that can account for the presence of
over-dispersion. All of these examples illustrate how the case-base
sampling framework in general, and the package \pkg{casebase} in
particular, allows the user to fit a broad and flexible family of
survival functions.

@@

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \pkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
that capture both the event type and time. Both approaches to modeling
yield user-friendly code. However, in terms of output, both approaches
differ significantly. Case-base sampling produces smooth hazards and
smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
models produce step-wise CIFs and never explicitly model the hazard
function. Qualitatively, we showed that by using splines in the linear
predictor, all three models yielded similar curves. However, the smooth
nature of the output of \pkg{casebase} provides a more intuitive
interpretation for consumers of these predictions. In Table
\ref{tab:compCBvsCox}, we provide a side-by-side comparison between the
Cox model and case-base sampling.

@@

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
would like to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools could then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
Again, we hope to include any resulting methodology as part of this
package.

@@

This report was generated on 2020-07-21 20:19:45 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.0 (2020-04-24)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.5
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.5.6               dotwhisker_0.5.0         
#>  [3] timereg_1.9.4             glue_1.4.1               
#>  [5] scales_1.1.0              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_0.8.5              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.0.2               ggplot2_3.3.0            
#> [13] tidyverse_1.3.0           lubridate_1.7.8          
#> [15] visreg_2.6.1              tibble_3.0.1             
#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.1-12          
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.1.0          glmnet_3.0-2             
#> [25] Matrix_1.2-18             data.table_1.12.8        
#> [27] cowplot_1.0.0             colorspace_1.4-1         
#> [29] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.0     
#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
#> [28] quantreg_5.55       tools_4.0.0         gtable_0.3.0       
#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.3.1        
#> [34] nlme_3.1-147        iterators_1.0.12    xfun_0.15          
#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
#> [40] muhaz_1.2.6.1       MASS_7.3-51.5       zoo_1.8-7          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
#> [58] rlang_0.4.6         pkgconfig_2.0.3     evaluate_0.14      
#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
#> [73] pillar_1.4.4        haven_2.2.0         foreign_0.8-78     
#> [76] mgcv_1.8-31         nnet_7.3-13         mstate_0.2.12      
#> [79] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
#> [82] jpeg_0.1-8.1        grid_4.0.0          readxl_1.3.1       
#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
#> [88] numDeriv_2016.8-1.1 stats4_4.0.0        munsell_0.5.0      
#> [91] viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}
